\chapter{Shared Database Server}


The labs in this course will use a number of technologies in both command-line and graphical environments.
Some resources will run on your local computer, others remotely on-site and others in the cloud.

Here you will setup your lab PC environment and familiarise yourself with key operations we will use each week.
The lab PCs have most of the software required already installed.


We will use a shared Linux server for some of our work in this module.
This will be accessed via SSH at:
\texttt{data.comp.dkit.ie}

This is accessible from eduroam WiFi and the lab network within DkIT but not outside of the campus.

\section{Signup}

\begin{enumerate}

\item SSH to the sign-up user at:
\begin{verbatim}
ssh signup@data.comp.dkit.ie
\end{verbatim}

\item When you see the prompt
\begin{verbatim}
signup@data.comp.dkit.ie's password: 
\end{verbatim}
  you should use the password given by the lecturer in class.

\item At the first prompt enter your desired username (should NOT be your student number).
  No spaces, all lowercase.

\item Copy your SSH key in full from:
\begin{verbatim}
~/.ssh/id_rsa.pub
\end{verbatim}
  
\item Paste your SSH key in full at the second prompt and press enter.

\item If your request is successful you'll get a confirmation message.
  
\end{enumerate}

\section{Logging in}

When you have received confirmation that your account has been created, you can log in by SSH to:
\begin{verbatim}
ssh yourname@data.comp.dkit.ie
\end{verbatim}
