\chapter{Introduction}

\section{Module introduction}

Module descriptor:\\
\url{https://courses.dkit.ie/index.cfm/page/module/moduleId/56922}


\section{Pre-module checks}

Consider the following questions: 

\begin{enumerate}
\item What database(s) if any have you used to date?
\item Are you familiar with command-line interfaces?
\item Have you used PowerShell before?
\item Have you used Linux or UNIX before?
\item Have you logged into a remote system using Secure Shell (or Telnet)?
\item Do you plan to use Lab computers or your own Laptop? 
\end{enumerate}


\section{Technology Choices}

\begin{itemize}
\item Primary programming language is Python.
\item PostgreSQL is our utilised RDBMS.
\item Primary code distribution will be the git version control system.
\item We will use local and remote resources. Very important in your future career!
\end{itemize}


\section{Textbooks}

A number of students have asked for suggestions about database texts.

The module descriptor lists:

\begin{quotation}
Connolly, Thomas \& Begg, Carolyn. (2015), Database Systems, 6th. Addison-Wesley, [ISBN: 1292061189]
\end{quotation}

which is one I regularly use.
In most cases the best resource by far is the PostgreSQL manual.
It is very comprehensive and presents the material in a suitably applied style:

\url{https://www.postgresql.org/docs/13/index.html}

Other texts include:

\begin{enumerate}
\item Mannino, Michael. (2014), Database Design, Application Development, and Administration, 6th. Chicago Business Press, [ISBN: 0983332428].
\item Bob Bryla. (2015), Oracle Database 12c DBA Handbook, McGrawHill, [ISBN: 9780071798785].
\item Craig S. Mullins. (2012), Database Administration: The Complete Guide to DBA Practices and Procedures, 2nd. Addison Wesley, [ISBN: 0321822943].
\item Coronel, C. ;Morris, S.. (2019), Database Systems: Design, Implementation, and Management, 13th ed.. Cengage Learning, [ISBN: 9781333762790].
\item DuBois, Paul. (2014), MySQL: Developers Library, 5th ed.. Pearson, [ISBN: 9780321833877].
\item Elmasri et al.. (2016), Fundamentals of Database Systems, 7th ed.. Pearson, [ISBN: 9780133970777].
\item Date, C.J.. (2015), SQL and Relational Theory, 3rd ed.. O'Reilly Media, [ISBN: 978149194117].
\end{enumerate}

